---
title: "ATAC-seq Differential Analysis — DiffBind + ChIPseeker"
author: "ATAC-seq Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: hide
    fig_width: 10
    fig_height: 7
params:
  samplesheet: "results/diffbind/samplesheet.csv"
  outdir: "results/diffbind"
  fdr: 0.05
  fold: 1.0
  method: "DESeq2"
  summits: 250
  min_overlap: 2
  threads: 8
---

```{r setup, include=FALSE}
# ─── Global options ───────────────────────────────────────────────────────
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "100%"
)

# ─── Load libraries ───────────────────────────────────────────────────────
suppressPackageStartupMessages({
  library(DiffBind)
  library(BiocParallel)
  library(ChIPseeker)
  library(TxDb.Hsapiens.UCSC.hg38.knownGene)
  library(org.Hs.eg.db)
  library(GenomicRanges)
  library(IRanges)
  library(ggplot2)
  library(dplyr)
  library(pheatmap)
  library(RColorBrewer)
  library(ggrepel)
  library(reshape2)
  library(gridExtra)
  library(knitr)
})

# ─── Parameters ───────────────────────────────────────────────────────────
fdr_cutoff   <- params$fdr
fold_cutoff  <- params$fold
method_name  <- params$method
n_threads    <- params$threads
summits_bp   <- params$summits
min_overlap  <- params$min_overlap

# Setup output directories
dir.create(file.path(params$outdir, "plots"),   showWarnings = FALSE, recursive = TRUE)
dir.create(file.path(params$outdir, "results"), showWarnings = FALSE, recursive = TRUE)

# Parallel workers
if (n_threads > 1) {
  register(MulticoreParam(workers = n_threads))
} else {
  register(SerialParam())
}

# Reference databases
txdb  <- TxDb.Hsapiens.UCSC.hg38.knownGene
annodb <- "org.Hs.eg.db"

method_flag <- if (method_name == "DESeq2") DBA_DESEQ2 else DBA_EDGER
```

---

# Overview

This report presents a complete **differential accessibility analysis** of ATAC-seq data using:
- **DiffBind**: Peak counting and normalization
- **DESeq2/edgeR**: Differential analysis
- **ChIPseeker**: Peak annotation

**Parameters:**
- FDR cutoff: `r fdr_cutoff`
- Fold-change cutoff: `r fold_cutoff`
- Method: `r method_name`
- Peak re-centering: ±`r summits_bp` bp

---

# 1. Exploratory Data Analysis (EDA)

## 1.1 Load Sample Information

```{r load-samplesheet}
# Load DiffBind object
cat("Loading samplesheet from:", params$samplesheet, "\n")
dba_obj <- dba(sampleSheet = params$samplesheet)
print(dba_obj)

# Extract sample metadata
sample_info <- dba_obj$samples
cat("\n")
print(sample_info)
```

### Sample Summary Table

```{r sample-summary-table, results='asis'}
sample_table <- data.frame(
  Sample = sample_info$SampleID,
  Condition = sample_info$Condition,
  Replicate = sample_info$Replicate,
  BAM = sample_info$bamReads,
  Reads = sample_info$bamReads
)
kable(sample_table, caption = "Sample Metadata Summary")
```

## 1.2 Count Reads in Peaks

```{r count-reads}
cat("Counting reads in peaks (SummarizeOverlaps) ...\n")
count_args <- list(DBA = dba_obj, bUseSummarizeOverlaps = TRUE)
if (summits_bp > 0) {
  count_args$summits <- summits_bp
  cat(sprintf("Re-centering peaks around summits ± %d bp\n", summits_bp))
}
dba_obj <- do.call(dba.count, count_args)
print(dba_obj)
```

## 1.3 Normalize Read Counts

```{r normalize}
cat(sprintf("Normalising using %s ...\n", method_name))
if (method_name == "DESeq2") {
  dba_obj <- dba.normalize(dba_obj, method = DBA_DESEQ2)
} else {
  dba_obj <- dba.normalize(dba_obj, method = DBA_EDGER)
}
print(dba_obj)
```

## 1.4 Descriptive Statistics

### Read Distribution by Sample and Condition

```{r read-distribution}
# Extract normalized counts
norm_matrix <- dba.peakset(dba_obj, bRetrieve = TRUE)

# Summary statistics
cat("Normalised read counts summary (per sample):\n")
col_sums <- colSums(assay(norm_matrix))
col_means <- colMeans(assay(norm_matrix))
stats_df <- data.frame(
  Sample = names(col_sums),
  Total_Reads = as.integer(col_sums),
  Mean_Per_Peak = round(col_means, 2),
  Median_Per_Peak = apply(assay(norm_matrix), 2, median),
  SD_Per_Peak = round(apply(assay(norm_matrix), 2, sd), 2)
)
print(stats_df)

# Add condition info
stats_df$Condition <- sample_info$Condition[match(stats_df$Sample, sample_info$SampleID)]
```

### Histogram of Read Counts

```{r read-count-histogram, fig.height=6, fig.width=10}
# Convert to long format for ggplot
count_long <- melt(as.matrix(assay(norm_matrix)))
colnames(count_long) <- c("Peak", "Sample", "Count")
count_long$Condition <- sample_info$Condition[match(count_long$Sample, sample_info$SampleID)]

# Plot
ggplot(count_long[count_long$Count > 0, ], aes(x = log10(Count + 1), fill = Condition)) +
  geom_histogram(bins = 50, alpha = 0.7) +
  facet_wrap(~Sample, scales = "free_y") +
  theme_bw() +
  labs(
    title = "Distribution of Normalised Peak Counts",
    x = "log10(Normalised Count + 1)",
    y = "Frequency"
  )
```

## 1.5 Peak Characteristics

### Number of Peaks

```{r peak-numbers}
# Get consensus peak set
consensus_peaks <- dba.peakset(dba_obj, bRetrieve = TRUE)
n_peaks <- nrow(consensus_peaks)

# Peaks per sample
peaks_per_sample <- apply(assay(norm_matrix) > 0, 2, sum)

cat(sprintf("Total consensus peaks: %d\n", n_peaks))
cat("\nPeaks detected per sample:\n")
print(peaks_per_sample)
```

### Peak Size Distribution

```{r peak-size-distribution}
# Extract peak widths
peak_widths <- width(consensus_peaks)
peak_stats <- data.frame(
  Min = min(peak_widths),
  Q1 = quantile(peak_widths, 0.25),
  Median = median(peak_widths),
  Mean = round(mean(peak_widths), 1),
  Q3 = quantile(peak_widths, 0.75),
  Max = max(peak_widths),
  SD = round(sd(peak_widths), 1)
)

kable(peak_stats, caption = "Peak Size Statistics (bp)", digits = 0)

# Histogram
ggplot(data.frame(Width = peak_widths), aes(x = Width)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  scale_x_log10() +
  theme_bw() +
  labs(
    title = "Distribution of Peak Widths",
    x = "Peak Width (bp, log10 scale)",
    y = "Frequency"
  )
```

---

# 2. Quality Control Plots

## 2.1 Correlation Heatmap

```{r qc-correlation, fig.height=8, fig.width=9}
pdf(file.path(params$outdir, "plots", "correlation_heatmap.pdf"), width = 11, height = 10)
dba.plotHeatmap(dba_obj, main = "Sample Pearson Correlation (Normalised Counts in Peaks)")
dev.off()

dba.plotHeatmap(dba_obj, main = "Sample Pearson Correlation (Normalised Counts in Peaks)")
```

**Interpretation:** Red = high correlation (similar peak patterns). Check that biological replicates cluster together.

## 2.2 Principal Component Analysis (PCA)

```{r qc-pca, fig.height=7, fig.width=8}
pdf(file.path(params$outdir, "plots", "pca_plot.pdf"), width = 8, height = 7)
dba.plotPCA(dba_obj, label = DBA_CONDITION, main = "PCA — All Samples (Normalised Counts)")
dev.off()

dba.plotPCA(dba_obj, label = DBA_CONDITION, main = "PCA — All Samples (Normalised Counts)")
```

**Interpretation:** Samples should cluster by condition. Great separation indicates strong condition effects.

## 2.3 Peak Overlap (Venn/UpSet)

```{r qc-venn, fig.height=7, fig.width=8, eval=TRUE}
conditions <- unique(sample_info$Condition)
if (length(conditions) <= 4) {
  pdf(file.path(params$outdir, "plots", "peak_overlap_venn.pdf"), width = 8, height = 7)
  tryCatch(
    dba.plotVenn(dba_obj, dba_obj$masks$All, main = "Peak Overlap Across All Samples"),
    error = function(e) message("Venn plot skipped: ", conditionMessage(e))
  )
  dev.off()
  
  dba.plotVenn(dba_obj, dba_obj$masks$All, main = "Peak Overlap Across All Samples")
} else {
  cat("Venn plot skipped: > 4 samples — too complex\n")
}
```

**Interpretation:** Overlapping circles show which peaks are shared across samples. High overlap suggests reproducible chromatin accessibility.

## 2.4 Read Count Distribution

```{r qc-boxplot, fig.height=6, fig.width=10}
pdf(file.path(params$outdir, "plots", "count_boxplot.pdf"), width = 10, height = 6)
dba.plotBox(dba_obj, main = "Normalised Read Counts in Peaks by Sample")
dev.off()

dba.plotBox(dba_obj, main = "Normalised Read Counts in Peaks by Sample")
```

---

# 3. Differential Accessibility Analysis

## 3.1 Define Contrasts

```{r set-contrasts}
cat(sprintf("Setting up pairwise contrasts (method: %s, min members: %d) ...\n", 
            method_name, min_overlap))
dba_obj <- dba.contrast(dba_obj, minMembers = min_overlap, method = method_flag)

contrast_tbl <- dba.show(dba_obj, bContrasts = TRUE)
cat("\nContrasts defined:\n")
print(contrast_tbl)
```

## 3.2 Run Differential Analysis

```{r differential-analysis}
cat(sprintf("Running differential analysis using %s ...\n", method_name))
dba_obj <- dba.analyze(dba_obj, method = method_flag, bParallel = FALSE)
```

## 3.3 Per-Contrast Results

```{r per-contrast-results}
all_results <- list()
contrast_tbl <- dba.show(dba_obj, bContrasts = TRUE)

for (i in seq_len(nrow(contrast_tbl))) {
  grp1 <- as.character(contrast_tbl[i, "Group"])
  grp2 <- as.character(contrast_tbl[i, "Group2"])
  cname <- paste0(grp1, "_vs_", grp2)
  
  cat(sprintf("\n--- Contrast %d: %s ---\n", i, cname))

  # Full results and significant results
  res_all <- tryCatch(
    dba.report(dba_obj, contrast = i, method = method_flag,
               th = 1, bNormalized = TRUE, bCalled = FALSE),
    error = function(e) { message("ERROR dba.report: ", e$message); NULL }
  )
  if (is.null(res_all)) next

  res_sig <- dba.report(dba_obj, contrast = i, method = method_flag,
                        th = fdr_cutoff, fold = fold_cutoff, bNormalized = TRUE)

  # Convert to data frame
  res_df <- as.data.frame(res_all)
  res_sig_df <- as.data.frame(res_sig)
  
  res_df$contrast <- cname
  res_sig_df$contrast <- cname
  res_df$significant <- res_df$FDR <= fdr_cutoff & abs(res_df$Fold) >= fold_cutoff

  all_results[[cname]] <- res_df

  # Write CSVs
  write.csv(res_df, 
            file = file.path(params$outdir, "results", paste0(cname, "_all_peaks.csv")),
            row.names = FALSE)
  write.csv(res_sig_df,
            file = file.path(params$outdir, "results", paste0(cname, "_significant_peaks.csv")),
            row.names = FALSE)

  cat(sprintf("  Total peaks: %d\n", nrow(res_df)))
  cat(sprintf("  Significant peaks (FDR < %g, |log2FC| ≥ %g): %d\n",
              fdr_cutoff, fold_cutoff, nrow(res_sig_df)))
  cat(sprintf("    - Up-regulated:   %d\n", sum(res_sig_df$Fold > 0)))
  cat(sprintf("    - Down-regulated: %d\n", sum(res_sig_df$Fold < 0)))
}
```

### Summary Table

```{r contrast-summary-table, results='asis'}
summary_list <- list()
for (cname in names(all_results)) {
  res_df <- all_results[[cname]]
  sig_peaks <- sum(res_df$significant)
  up_peaks <- sum(res_df$significant & res_df$Fold > 0)
  dn_peaks <- sum(res_df$significant & res_df$Fold < 0)
  
  summary_list[[cname]] <- data.frame(
    Contrast = cname,
    Total_Peaks = nrow(res_df),
    Significant_Peaks = sig_peaks,
    Up_regulated = up_peaks,
    Down_regulated = dn_peaks
  )
}
summary_df <- do.call(rbind, summary_list)
rownames(summary_df) <- NULL
kable(summary_df, caption = sprintf("Differential peaks (FDR<%g, |log2FC|≥%g)", 
                                     fdr_cutoff, fold_cutoff))
```

---

# 4. Volcano and MA Plots

```{r volcano-plots, fig.height=6, fig.width=7}
thresh_y <- -log10(fdr_cutoff + 1e-300)

for (cname in names(all_results)) {
  res_df <- all_results[[cname]]
  
  # Volcano plot
  p_volcano <- ggplot(res_df, aes(x = Fold, y = -log10(FDR + 1e-300))) +
    geom_point(aes(colour = significant), alpha = 0.5, size = 1) +
    scale_colour_manual(
      values = c("TRUE" = "#c0392b", "FALSE" = "grey60"),
      labels = c("Not sig.",
                 sprintf("FDR<%g  |log2FC|≥%g", fdr_cutoff, fold_cutoff)),
      name = NULL
    ) +
    geom_vline(xintercept = c(-fold_cutoff, fold_cutoff),
               linetype = "dashed", colour = "navy", linewidth = 0.3) +
    geom_hline(yintercept = thresh_y,
               linetype = "dashed", colour = "navy", linewidth = 0.3) +
    labs(
      title = sprintf("Volcano — %s", cname),
      x = expression(log[2]~Fold~Change),
      y = expression(-log[10](FDR))
    ) +
    theme_bw(base_size = 11)
  
  print(p_volcano)
  
  ggsave(file.path(params$outdir, "plots", paste0(cname, "_volcano.pdf")),
         p_volcano, width = 7, height = 6)
}
```

```{r ma-plots, fig.height=5, fig.width=7}
for (i in seq_len(nrow(contrast_tbl))) {
  pdf(file.path(params$outdir, "plots", 
                paste0(names(all_results)[i], "_MA.pdf")), 
      width = 7, height = 5)
  dba.plotMA(dba_obj, contrast = i, method = method_flag, th = fdr_cutoff)
  dev.off()
}

cat("MA plots saved to", file.path(params$outdir, "plots"), "\n")
```

---

# 5. Peak Annotation

```{r peak-annotation}
cat("\nAnnotating significant peaks with ChIPseeker ...\n")

for (cname in names(all_results)) {
  res_df <- all_results[[cname]]
  sig_df <- res_df[res_df$significant, , drop = FALSE]
  
  if (nrow(sig_df) < 5) {
    cat(sprintf("  %s: only %d significant peaks — skipping annotation\n",
                cname, nrow(sig_df)))
    next
  }
  
  cat(sprintf("  %s: annotating %d peaks ...\n", cname, nrow(sig_df)))
  
  # Convert to GRanges
  sig_gr <- GRanges(
    seqnames = sig_df$seqnames,
    ranges = IRanges(sig_df$start, sig_df$end),
    strand = "*"
  )
  
  # Annotate
  anno <- tryCatch(
    annotatePeak(sig_gr, tssRegion = c(-3000, 3000),
                 TxDb = txdb, annoDb = annodb, verbose = FALSE),
    error = function(e) { message("  ERROR: ", e$message); NULL }
  )
  if (is.null(anno)) next
  
  anno_df <- as.data.frame(anno)
  write.csv(anno_df,
            file = file.path(params$outdir, "results",
                             paste0(cname, "_significant_annotated.csv")),
            row.names = FALSE)
  
  # Annotation bar plot
  pdf(file.path(params$outdir, "plots", paste0(cname, "_annotation_bar.pdf")),
      width = 8, height = 5)
  plotAnnoBar(anno, title = sprintf("Peak Annotation — %s", cname))
  dev.off()
  
  # TSS distance plot
  pdf(file.path(params$outdir, "plots", paste0(cname, "_tss_distance.pdf")),
      width = 8, height = 5)
  plotDistToTSS(anno, title = sprintf("Distance to TSS — %s", cname))
  dev.off()
}
```

---

# 6. Save Session Data

```{r save-rdata}
# Combine all results
all_results_df <- do.call(rbind, all_results)
write.csv(all_results_df,
          file = file.path(params$outdir, "differential_peaks_all_contrasts.csv"),
          row.names = FALSE)

# Save RData for future interactive analysis
save(dba_obj, all_results, file = file.path(params$outdir, "diffbind_analysis.RData"))

cat("\n✓ RData saved to", file.path(params$outdir, "diffbind_analysis.RData"), "\n")
cat("\nAll outputs saved to:", params$outdir, "\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```

